Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-05-12 Manuscript/HI_Model_ms20210405.Rmd
updated manuscript files

```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = FALSE)
```
\centering
Paul C. Selmants^1^, Benjamin M. Sleeter^2^,  Jinxun Liu^1^, Tamara S. Wilson^1^,

@@

**Affiliations:**  
^1^U.S. Geological Survey, Moffett Field, CA, USA     
^2^U.S. Geological Survey, Seattle, WA, USA    
^3^University of Hawaiʻi at Mānoa, Honolulu, HI, USA  
^4^East-West Center, Honolulu, HI, USA     
^5^Clark University, Worcester, MA, USA   
^6^Arizona State University, Tempe, AZ, USA

@@

## Abstract
The State of Hawaiʻi passed legislation to be carbon neutral by 2045, a goal that will partly depend on carbon sequestration by terrestrial ecosystems. However, there is considerable uncertainty surrounding the future direction and magnitude of the land carbon sink in the Hawaiian Islands. We used simulation modeling to assess how projected future changes in climate and land use will influence ecosystem carbon balance in the Hawaiian Islands under all combinations of two radiative forcing scenarios (RCP 4.5 and RCP 8.5) and two land use scenarios (low and high) over a 90-year timespan from 2010-2100. Collectively, terrestrial ecosystems of the Hawaiian Islands acted as a net carbon sink under low radiative forcing (RCP 4.5) for the entire 90-year simulation period, with low land use change further enhancing carbon sink strength. In contrast, Hawaiian terrestrial ecosystems transitioned from a net sink to a net source of CO~2~ to the atmosphere under high radiative forcing (RCP 8.5), with high land use accelerating this transition and exacerbating net carbon loss. A sensitivity test of the CO~2~ fertilization effect on plant productivity revealed it to be a major source of uncertainty in projections of ecosystem carbon balance. Greater mechanistic understanding of plant productivity responses to rising atmospheric CO~2~ will be essential to realistically constrain simulation models used to evaluate the effectiveness of ecosystem-based climate mitigation strategies.

## 1. Introduction
Terrestrial ecosystems are a major sink for atmospheric carbon dioxide (CO~2~), removing ~30% of human emissions on an annual basis and reducing the rate of increase in atmospheric CO~2~ [@keenan_terrestrial_2018; @friedlingstein_global_2019]. There is increasing recognition among policymakers that natural and agricultural ecosystems can contribute to climate mitigation, which has given rise to the popularity of “natural climate solutions” [@cameron_ecosystem_2017]. Defined as conservation and land management efforts aimed at enhancing ecosystem carbon storage [@griscom_natural_2017], natural climate solutions are appealing because they are seen as cost-efficient and readily available [@averchenkova_climate_2017; @cameron_ecosystem_2017; @fargione_natural_2018]. However, effective implementation is complicated by the uncertainty surrounding the future direction and magnitude of the land carbon sink, especially at the regional scale. Despite this uncertainty, evidence indicates that both interannual and long-term variability in carbon uptake by land ecosystems is driven primarily by fluctuations in climate, land use, and land cover change [@ahlstrom_dominant_2015; @prestele_current_2017; @friedlingstein_global_2019]. Incorporating the interactive effects of land use and climate into spatially explicit future projections of ecosystem carbon balance could therefore provide a reference point to evaluate the effectiveness of land-based mitigation. Although a complex challenge, the growing number of sub-national jurisdictions that plan to incorporate land-based mitigation strategies into their emissions reduction efforts would benefit from understanding how future land use and climate-biosphere feedbacks will affect ecosystem carbon balance in their respective regions [@sleeter_effects_2019].

@@

### *4.2 Comparison to other studies*
There are few estimates of contemporary ecosystem carbon balance for the main Hawaiian Islands, and even fewer model projections of future ecosystem carbon balance in response to climate and land use change.  Our mean annual NBP estimate of 0.64 TgC y^-1^ for the period 2011-2020 agrees well with a recent State of Hawaiʻi Greenhouse Gas Inventory report, which estimated an annual net carbon sink of 0.66 Tg C in 2015 from agriculture, forestry, and other land uses [@hawaii_inventory_2019]. In contrast, our NBP estimate for the past decade was ~ 88% higher than a previous statewide LUCAS model estimate covering the same time period [0.341 Tg C y^-1^; @selmants_baseline_2017]. This discrepancy was likely driven by modifications in how we calculated NPP, soil R~h~, and soil aquatic carbon loss compared to previous versions of the LUCAS model, as well our model's finer spatial resolution [@selmants_baseline_2017; @daniel_integrating_2018]. Previous versions of a Hawaiʻi LUCAS model were run at 1-km spatial resolution and simulation cells within each unique combination of moisture zone and state type all had the same mean IBIS-derived NPP value applied to them at the beginning of the simulation period. In contrast, our NPP estimates at 250-m spatial resolution were adjusted on a cell-by-cell basis using Hawaiʻi-specific climate data as described in section 2.3. As a result, our statewide NPP estimates from 2011-2020 were 9.5% lower on average than previous LUCAS model estimates for Hawaiʻi during the same time period [@selmants_baseline_2017], likely because of the greater influence of more arid simulation cells. Soil carbon losses via R~h~, leaching, and overland flow in previous versions of the LUCAS model were calculated as the ratio of the IBIS-derived flux to the size of the originating carbon stock, in this case soil organic carbon to 1-m depth [@daniel_integrating_2018]. Here we calculated soil R~h~ and aquatic carbon losses as the ratio of the mean IBIS-derived flux to the microclimate-adjusted NPP value of each simulation cell, which is a more realistic driver than stock size [@jackson_ecology_2017]. Compared to previous Hawaiʻi LUCAS model estimates [@selmants_baseline_2017], soil R~h~ and aquatic carbon losses from 2011-2020 were reduced by an average of 15% and 21%, respectively, which widened the gap between between carbon gain (NPP) and carbon losses and accounted for the overall increase in NBP estimates for this time period.

@@

Our model does not currently differentiate between forests dominated by native versus non-native tree species, which could influence estimates of ecosystem carbon balance. Native forests in the Hawaiian Islands are dominated by ʻōhiʻa (_Metrosideros polymorpha_ Gauditch), an endemic foundational tree species found across a broad range of climatic and edaphic conditions [@ziegler_hawaiian_2002]. Beginning in 2010, a fungal disease termed Rapid ʻŌhiʻa Death (ROD) caused by two _Ceratocystus_ spp. has emerged and caused widespread mortality to mature ʻōhiʻa trees across a range of size classes, primarily on Hawaiʻi Island [@mortenson_assessing_2016; @fortini_evolving_2019]. The distribution and potential range of this emerging threat to Hawaiʻi's dominant native tree species have only recently been mapped [@vaughn_approach_2018; @fortini_evolving_2019], but the pulse of newly dead organic matter and reduction in photosynthetic capacity induced by widespread tree mortality could significantly alter ecosystem carbon balance over the long-term [@sleeter_effects_2019]. ROD-affected forests could also undergo a replacement of canopy dominant stress-tolerating ʻōhiʻa trees by non-native tree species with faster relative growth rates, lower wood density, and faster tissue turnover, potentially altering the long-term trajectory of carbon cycling in Hawaiian forest ecosystems. New model projections for Hawaiʻi that incorporate ROD spread rates and forest restoration scenarios will therefore require differentiation among forest ecosystems dominated by native and non-native tree species.

>>>>

```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = FALSE)
```
\centering
Paul C. Selmants^1,7^, Benjamin M. Sleeter^2^,  Jinxun Liu^1^, Tamara S. Wilson^1^,

@@

**Affiliations:**  
^1^U.S. Geological Survey, Moffett Field, CA, USA     
^2^U.S. Geological Survey, Seattle, WA, USA    
^3^University of Hawaiʻi at Mānoa, Honolulu, HI, USA  
^4^East-West Center, Honolulu, HI, USA 
^5^Clark University, Worcester, MA, USA   
^6^Arizona State University, Tempe, AZ, USA    
^7^Author to whom correspondence should be addressed

@@

## Abstract
The State of Hawaiʻi passed legislation to be carbon neutral by 2045, a goal that will partly depend on carbon sequestration by terrestrial ecosystems. However, there is considerable uncertainty surrounding the future direction and magnitude of the land carbon sink in the Hawaiian Islands. We used simulation modeling to assess how projected future changes in climate and land use will influence ecosystem carbon balance in the Hawaiian Islands under all combinations of two radiative forcing scenarios (RCPs 4.5 and 8.5) and two land use scenarios (low and high) over a 90-year timespan from 2010-2100. Collectively, terrestrial ecosystems of the Hawaiian Islands acted as a net carbon sink under low radiative forcing (RCP 4.5) for the entire 90-year simulation period, with low land use change further enhancing carbon sink strength. In contrast, Hawaiian terrestrial ecosystems transitioned from a net sink to a net source of CO~2~ to the atmosphere under high radiative forcing (RCP 8.5), with high land use accelerating this transition and exacerbating net carbon loss. A sensitivity test of the CO~2~ fertilization effect on plant productivity revealed it to be a major source of uncertainty in projections of ecosystem carbon balance. Greater mechanistic understanding of plant productivity responses to rising atmospheric CO~2~ will be essential to realistically constrain simulation models used to evaluate the effectiveness of ecosystem-based climate mitigation strategies.

## 1. Introduction
Terrestrial ecosystems are a major sink for atmospheric carbon dioxide (CO~2~), removing ~30% of human emissions on an annual basis and reducing the rate of increase in atmospheric CO~2~ [@keenan_terrestrial_2018; @friedlingstein_global_2019]. There is increasing recognition among policymakers that natural and agricultural ecosystems can contribute to climate mitigation, which has given rise to the popularity of “natural carbon solutions” [@cameron_ecosystem_2017]. Defined as conservation and land management efforts aimed at enhancing ecosystem carbon storage [@griscom_natural_2017], natural climate solutions are appealing because they are seen as cost-efficient and readily available [@averchenkova_climate_2017; @cameron_ecosystem_2017; @fargione_natural_2018]. However, effective implementation is complicated by the uncertainty surrounding the future direction and magnitude of the land carbon sink, especially at the regional scale. Despite this uncertainty, evidence indicates that both interannual and long-term variability in carbon uptake by land ecosystems is driven primarily by fluctuations in climate, land use, and land cover change [@ahlstrom_dominant_2015; @prestele_current_2017; @friedlingstein_global_2019]. Incorporating the interactive effects of land use and climate into spatially explicit future projections of ecosystem carbon balance could therefore provide a reference point to evaluate the effectiveness of land-based mitigation. Although a complex challenge, the growing number of sub-national jurisdictions that plan to incorporate land-based mitigation strategies into their emissions reduction efforts would benefit from understanding how future land use and climate-biosphere feedbacks will affect ecosystem carbon balance in their respective regions [@sleeter_effects_2019].

@@

### *4.2 Comparison to other studies*
There are few estimates of contemporary ecosystem carbon balance for the main Hawaiian Islands, and even fewer model projections of future ecosystem carbon balance in response to climate and land use change.  Our mean annual NBP estimate of 0.64 TgC y^-1^ for the period 2011-2020 agrees well with a recent State of Hawaiʻi Greenhouse Gas Inventory report, which estimated an annual net carbon sink of 0.66 Tg C in 2015 from agriculture, forestry, and other land uses [@hawaii_inventory_2019]. In contrast, our NBP estimate for the past decade was ~ 88% higher than a previous statewide LUCAS model estimate covering the same time period [0.341 Tg C y^-1^; @selmants_baseline_2017]. This discrepancy was likely driven by modifications in how we calculated NPP, soil R~h~, and soil aquatic carbon loss compared to previous versions of the LUCAS model, as well our modelʻs finer spatial resolution [@selmants_baseline_2017; @daniel_integrating_2018]. Previous versions of a Hawaiʻi LUCAS model were run at 1-km spatial resolution and simulation cells within each unique combination of moisture zone and state type all had the same mean IBIS-derived NPP value applied to them at the beginning of the simulation period. In contrast, our NPP estimates at 250-m spatial resolution were adjusted on a cell-by-cell basis using Hawaiʻi-specific climate data as described in section 2.3. As a result, our statewide NPP estimates from 2011-2020 were 9.5% lower on average than previous LUCAS model estimates for Hawaiʻi during the same time period [@selmants_baseline_2017], likely because of the greater influence of more arid simulation cells. Soil carbon losses via R~h~, leaching, and overland flow in previous versions of the LUCAS model were calculated as the ratio of the IBIS-derived flux to the size of the originating carbon stock, in this case soil organic carbon to 1-m depth [@daniel_integrating_2018]. Here we calculated soil R~h~ and aquatic carbon losses as the ratio of the mean IBIS-derived flux to the microclimate-adjusted NPP value of each simulation cell, which is a more realistic driver than stock size [@jackson_ecology_2017]. Compared to previous Hawaiʻi LUCAS model estimates [@selmants_baseline_2017], soil R~h~ and aquatic carbon losses from 2011-2020 were reduced by an average of 15% and 21%, respectively, which widened the gap between between carbon gain (NPP) and carbon losses and accounted for the overall increase in NBP estimates for this time period.

@@

Our model does not currently differentiate between forests dominated by native versus non-native tree species, which could influence estimates of ecosystem carbon balance. Native forests in the Hawaiian Islands are dominated by ʻōhiʻa (_Metrosideros polymorpha_ Gauditch), an endemic foundational tree species found across a broad range of climatic and edaphic conditions [@ziegler_hawaiian_2002]. Beginning in 2010, a fungal disease termed Rapid ʻŌhiʻa Death (ROD) caused by two _Ceratocystus_ spp. has emerged and caused widespread mortality to mature ʻōhiʻa trees across a range of size classes, primarily on Hawaiʻi Island [@mortenson_assessing_2016; @fortini_evolving_2019]. The distribution and potential range of this emerging threat to Hawaiʻi's dominant native tree species have only recently been mapped [@vaughn_approach_2018; @fortini_evolving_2019], but the pulse of newly dead organic matter and reduction in photosynthetic capacity induced by widespread tree mortality could significantly alter ecosystem carbon balance over the long-term [@sleeter_effects_2019]. ROD-affected forests could also undergo a replacement of canopy dominant stress-tolerating ʻōhiʻa trees by non-native tree species with faster relative growth rates, lower wood density, and faster tissue turnover, potentially alter the long-term trajectory of forest carbon cycling in Hawaiʻi. New model projections for Hawaiʻi that incorporate ROD spread rates and forest restoration scenarios will therefore require differentiation among forest ecosystems dominated by native and non-native tree species.
